Bayesian inference of relative fitness on high-throughput pooled competition assays

### 2023-12-20 doc/_discussion.qmd
Add explanation of fitness model dynamics and future directions

It is important to highlight some of the consequences of the general
experimental design and the implicit assumptions within the proposed statistical
model to analyze the resulting data. First, the composition of the population is
such that the initial fraction of the population occupied by the barcoded
genotypes is small---usually >90\% of the initial population is the non-labeled
reference strain. This constraint is important as the fitness model used to fit
the time series data assumes that the tracked frequencies are $\ll 1$. Second,
when computing log frequency ratios, we can run into the issue of dividing by
zero. This is a common problem when dealing with molecular count data
[@lovell2020]. Our model gets around this issue by assuming that the frequency
of any barcode cannot be, but still can get arbitrarily close to, zero.
Therefore, we implicitly assume that no lineage goes extinct during the
experiment. Moreover, the statistical model directly accounts for the
uncertainty associated with having zero barcode counts, increasing the
corresponding uncertainty. Third, the models presented in this paper require the
existence of a labeled sub-population of barcoded reference strains. These
barcodes help determine the fitness baseline, as every fitness is quantified
with respect to this reference genotype. This experimental design constraint
facilitates the inference of the population mean fitness since most of the
culture---the unlabeled reference genotype---is not tracked. Finally, the
presented statistical model assumes that relative fitness is solely a constant
of the environment and the genotype. Future directions of this work could extend
the fitness model to properly analyze data with time-varying or
frequency-dependent fitness values.

>>>>

It is important to highlight some of the consequences of the general
experimental design and the implicit assumptions within the proposed statistical
model to analyze the resulting data. First, the composition of the population is
such that the initial fraction of the population occupied by the barcoded
genotypes is small---usually >90\% of the initial population is the non-labeled
reference strain. This constraint is important as the fitness model used to fit
the time series data assumes that the tracked frequencies are $\ll 1$. Second,
when computing log frequency ratios, we can run into the issue of dividing by
zero. This is a common problem when dealing with molecular count data
[@lovell2020]. Our model gets around this issue by assuming that the frequency
of any barcode cannot be, but still can get arbitrarily close to, zero.
Therefore, we implicitly assume that no lineage goes extinct during the
experiment. Moreover, the statistical model directly accounts for the
uncertainty associated with having zero barcode counts, increasing the
corresponding uncertainty. Third, the models presented in this paper require the
existence of a labeled sub-population of barcoded reference strains. These
barcodes help determine the fitness baseline, as every fitness is quantified
with respect to this reference genotype. This experimental design constraint
facilitates the inference of the population mean fitness since most of the
culture---the unlabeled reference genotype---is not tracked. 
<span style="color: #912E27;">
Fourth, the fitness model in @eq-fitness does not assume any functional form for
the cellular growth dynamics. This description of the changes in the relative
frequency of the genotypes can approximate the dynamics for multiple growth
forms. For example, in @sec-fitseq2, in the supplementary materials, we apply
our inference pipeline to simulations like those presented in [@li2023] with
non-logistic growth models. Moreover, @sec-kinsler, in the supplementary
materials, reanalyzes experimental data from [@kinsler2020], where previous work
showed that the fitness advantage from some of the mutants results from a
shorter lag phase at the beginning of the growth cycle [@li2018]. Thus, our
method's inferred relative fitness is a coarse-grained quantity that integrates
the differences in fitness across the entire growth cycle as long as frequencies
and frequency changes are small enough to be approximated by @eq-fitness.
</span>
Finally, the presented statistical model assumes that relative fitness is solely
a constant of the environment and the genotype. Future directions of this work
could extend the fitness model to properly analyze data with time-varying or
frequency-dependent fitness values.
